A Bayesian search to find \\high-mass black holes in LIGO data

### 2021-01-05 main.tex
Updates from Overleaf
incorporating rory's suggestions

\documentclass[%
 %reprint,
%superscriptaddress,
%groupedaddress,
%unsortedaddress,
%runinaddress,
%frontmatterverbose, 
preprint,
%preprintnumbers,
%nofootinbib,
%nobibnotes,
%bibnotes,
 amsmath,amssymb,
 aps,
%pra,
%prb,
%rmp,
%prstab,
%prstper,
%floatfix,
]{revtex4}

@@

\newcommand{\prior}{{\sc prior}\xspace}
\newcommand{\bilby}{{\sc \href{https://lscsoft.docs.ligo.org/bilby/}{\texttt{Bilby}}}\xspace}
\newcommand{\bilbypipe}{{\sc bilby\_pipe}\xspace}
\newcommand{\pbilby}{{\sc pBilby}\xspace}
\newcommand{\dynesty}{{\sc dynesty}\xspace}
\newcommand{\cpnest}{{\sc cpnest}\xspace}
\newcommand{\ptemcee}{{\sc ptemcee}\xspace}
\newcommand{\gwpy}{{\sc \href{https://gwpy.github.io/}{\texttt{GWpy}}}\xspace}
\newcommand{\imrphenomp}{{\sc IMRPhenomPv2}\xspace}
\newcommand{\seob}{{\sc SEOBNRv4PHM}\xspace}
\newcommand{\pycbc}{{\sc \href{https://pycbc.org/}{\texttt{PyCBC}}}\xspace}
\newcommand{\bcr}{{\sc BCR}\xspace}
\newcommand{\psd}{{\sc $\sigma^2_i$}\xspace}
\newcommand{\pastro}{{\sc $p_\text{astro}$}\xspace}
\newcommand{\msun}{{\sc M${}_\odot$}\xspace}
\newcommand{\template}{{\sc $\mu(\vec{\theta})$}\xspace}
\newcommand{\parameters}{{\sc $\vec{\theta}$}\xspace}

@@

\preprint{APS/123-QED}

\title{A Bayesian search to find \\high-mass black holes in LIGO data}% Force line breaks with \\

@@

The detection of high mass black holes ( $>100$ \msun) will shed light on the formation of supermassive black holes and thus galaxy formation. Although LIGO is sensitive to the merger of binary black holes with total masses up to 500 \msun, the largest total mass detected so far is approximately $80$ \msun. A possible explanation for the absence of high mass events may be due to their misclassification as short-duration instrumental noise transients. Short-duration instrumental transients mimic the short-duration gravitational-wave signals from high-mass binary black hole mergers. Here we demonstrate that a new search method utilising Bayesian inference could be a more sensitive tool for detecting high-mass binary black hole mergers as compared to traditional match-filtering. We have applied this technique on the high-mass triggers during LIGO's second observing run to investigate the possibility of discovering new gravitational-wave signals from an entirely new class of high mass black hole binaries. \rs{add sentence describing if we see any new events; add a sentence on our analysis of IAS events}

@@

%\tableofcontents

%%%%%%---SECTIONS---%%%%%%%%%%%%
\section{\label{sec:Introduction}Introduction}
\av{Need to edit in light of S190521G paper}
Since the 1970s, there has been an accumulation of evidence for stellar-mass and supermassive black holes. In 2019, the Event Horizon Telescope provided the first visual evidence of the supermassive black hole M87  \cite{m87photo}.  As of January 2020, LIGO, Virgo and IAS \rs{we should think carefully about how to describe the IAS events here; I think our conclusion is that they're not real} have confirmed more than a dozen binary black hole systems \cite{GWTC1, IAS0, IAS1, IAS2, pycbc_ogc_2}. These various discoveries have firmly established the existence of stellar-mass black holes, supermassive black holes and binary black hole systems.  Interestingly, there is still no direct evidence for intermediate-mass black holes, the black holes that lie in between stellar-mass and supermassive black hole systems with masses between $10^2-10^6$ \msun. Thus, the existence of intermediate-mass black holes is still speculative. \\

If intermediate-mass black holes are present, the gravitational waves emitted from the merger of a binary black hole system with at least one intermediate-mass black hole (up to a mass of $400$ \msun) will be detectable by the ground-based gravitational-wave network. Gravitational waves from such systems should occur at a rate of $0-10\text{yr}^{-1}$ \cite{fregeau2006imbhbRatePrediction, mandel2008rates,rodriguez2015bbhRatePredictions}.  However, even after conducting a targeted matched-filter based search for gravitational waves from intermediate-mass black holes the largest total mass detected so far is approximately $80$ \msun \cite{imbhbSearch2014, ligo_imbh_search, abbott2019gwtc}. A possible explanation for the absence of intermediate-mass events may be due to their misclassification as short-duration instrumental noise transients known as glitches \cite{glitch_in_fifth_ligo_run, bayeswave, improving_dq_in_early_runs, ligo_glitch_gw150914, pycbc_short_duration_transients, pe_with_glitch, blip_glitches}. These glitches can mimic astrophysical signals and hence decrease the significance of true gravitational wave events. \\

One method to account for glitches while searching for gravitational waves from coalescing compact binaries is by utilising an astrophysical Bayesian odds  \cite{bci, BCR1, BCR2, bcr_gw151216, bayesian_odds}. A true Bayesian odds calculated without using bootstrap techniques can provide events with a more accurate significance that is agnostic to a specific search strategy \cite{BCR2, bcr_gw151216,  bayesian_odds}. Additionally, a Bayesian odds can include more information than is included in current matched filter searches such as if the gravitational wave event's signal is coherent amongst the network of detectors, if the binary system that created the gravitational waves was precessing, and if the gravitational wave signal contains higher-order modes.  It is because Bayesian methods can incorporate all this physical information about a gravitational wave signal that the LIGO Scientific Collaboration uses these methods to determine the source parameters of gravitational wave events \cite{abbott2016ligo, abbott2019gwtc}. This paper demonstrates that the power of Bayesian methods used in parameter estimation can also successfully be used to discriminate between coherent gravitational-wave signals, incoherent glitches, and Gaussian noise in the form of a Bayesian search. \\

In this paper we utilise a Bayesian method, called the \bcr \cite{BCR1}, to search for the significant gravitational wave signals from high-mass (systems with total masses in the range of $55-500$ \msun) compact binary coalescences in the detector data recorded during aLIGO's second observing run (O2).  We find that (a) our search does not identify any unreported stellar mass or intermediate mass black holes; (b) high-mass events reported in the GWTC-1, including GW170729 (an event with disputed $p_\text{astro}$ amongst various search pipelines) have high low significance; and that (c) high-mass events detected from some independent groups have low significance. \rs{explicitly state that they're the IAS events}\\

@@

\section{Method\label{sec:method}}
\rs{I think you need a synopsis of this section before going into the details; explain that one could in principle apply bayesian methods only, but we don't do that here. Instead, we take triggers from pycbc and use bayesian evidences as a ranking statistic etc...} 
\subsection{Triggers for Analysis}
The LIGO Scientific collaboration runs a number of search pipelines that target gravitational waves from compact binary mergers such as the GstLAL, MBTA, SPIIR and PyCBC searches \cite{abbott2019gwtc}.\\

At its core, the \pycbc \cite{pycbc_code} search performs a matched-filter search for binary merger signals by comparing the data against templates from a gravitational-wave template-bank \cite{pycbc_og6}. \sout{\pycbc calculates the matched-filter for several different coalescence times and uses numerous templates from the template bank for each section of data being analysed.} The output of \pycbc's search is a signal-to-noise ratio $\rho$ like statistic, called the \pycbc ranking statistic $\rho_\text{PC}$. This statistic is a function of the matched-filter $\rho$, signal-based veto values, intrinsic and extrinsic properties of the apparent signal and other information \cite{pycbc_og6}. Whenever a local maximum of $\rho_\text{PC} > \rho_\text{T}$, where $\rho_\text{T}$ is some threshold value, the search pipeline produces a single-detector trigger associated with the detector and time where the data was obtained.\\

For \pycbc to consider a trigger to be a \textbf{candidate trigger}, a trigger from astrophysical origins, the trigger must be observed between detectors with the same template and a time of arrival difference less than the gravitational-wave travel time between detectors \cite{pycbc_og6}. To test its search, \pycbc also conducts searches for \textbf{simulated triggers}, artificial triggers manufactured by injecting signals into the detector data. These simulated signal studies provides \pycbc with metrics on its search's sensitivity. Finally, to quantify the statistical significance of candidate events, \pycbc artificially constructs a \textbf{background trigger} set to compare the candidate events against. These background triggers are signal-free events \rs{not quite: they contain signals, just not coherent ones}, constructed by applying relative offsets, or time-shifts, between the data of different detectors \cite{pycbc_og6}. Note that the time-shifts to generate the background triggers are greater than the gravitational-wave travel time between detectors to ensure that the background triggers are not of astrophysical origins. \\

In our work, we demonstrate that the \bcr can be used in a similar way as $\rho_\text{PC}$, as a ranking statistic, to measure the statistical significance of candidate triggers. Before we discuss how we use the \bcr as a measure of significance, we introduce the method to calculate the \bcr in the following section.\\

@@

Bayes theorem states that the probability $P(\mathcal{H}|d)$ for a hypothesis $\mathcal{H}$ and data $d$ is
\begin{equation}
    P(\mathcal{H}|d) = \frac{\mathcal{L}(d|\mathcal{H})\pi(\mathcal{H})}{Z}\ , 
\end{equation}
\rs{see comments from last week on replacing L wiht Z etc...}
where $\mathcal{L}(d|\mathcal{H})$ is the likelihood of the data given
the hypothesis,  $\pi(\mathcal{H})$ is the prior probability of the hypothesis, and finally, $Z = \sum \mathcal{L}(d|\mathcal{H})  \pi(\mathcal{H})$ is the marginalised likelihood called the evidence \av{Rory suggested we use $Z(d|H)$ instead of $L(D|H)$}.  To compare two hypotheses $\mathcal{H}_A$ and $\mathcal{H}_B$ with the Bayes theorem one can calculate an odds ratio
\begin{equation}
    \mathcal{O}^A_B = \frac{Z^A\ \pi(\mathcal{H}_A)}{Z^B\ \pi(\mathcal{H}_B)}\ ,
\end{equation}
where $Z^A$ and $Z^B$ are the evidences for the $\mathcal{H}_A$ and $\mathcal{H}_B$ hypotheses. The odds ratio can tell us which of the two hypotheses is more likely. For example, if $\mathcal{O}^A_B >> 1$, then this odds ratio indicates that the $\mathcal{H}_A$ is much more likely than $\mathcal{H}_B$. \\

The \bcr is a Bayesian odds ratio like the above, of a coherent signal hypotheses $\mathcal{H}_S$ and an incoherent instrumental feature hypothesis $\mathcal{H}_I$ for a network of $D$ detectors. $\mathcal{H_I}$ states that each detector $i$ has either pure Gaussian noise $\mathcal{H}_N$ or a glitch $\mathcal{H}_G$. Following \citet{BCR1}, the \bcr is given by
\begin{equation}
\label{eq:bcr}
\text{BCR} = \frac{\alpha Z^S}{\prod\limits^D_{i=1} \ [\beta Z^G_i + (1-\beta)Z^N_i]}\ ,
\end{equation}
where $Z^S$, $Z^G_i$ and $Z^N_i$ are the Bayesian evidences (marginalised likelihoods) for $\mathcal{H}_S$, $\mathcal{H}_N$, and $\mathcal{H}_G$. $\alpha$ and $\beta$, are the prior odds for obtaining a signal $\alpha=P(\mathcal{H}_S)/P(\mathcal{H}_I)$ and the prior odds for obtaining a glitch $\beta=P(\mathcal{H}_G)/P(\mathcal{H}_I)$. As the rate of signal and glitches are unknown, these priors $\alpha$ and $\beta$ are tuned to maximise the \bcr distributions for background data (signal-free data) and simulated signals \cite{BCR1}.  \\

\subsection{Bayesian Evidence Evaluation}
\subsubsection{Noise Model}
We assume that each detector's noise is Gaussian and stationary over the period being analysed \cite{ligo_psd}. In practice, we estimate the noise variance using the noise power spectral density (PSD)  \psd of the data. \rs{NOTE: $\sigma_i^2$ is the noise variance, not the PSD. If you compare the likelihood in, e.g., veitch and vecchio to yours you can see how to relate the variance to the PSD and vice versa}. Using the \psd, for each data segment $d_i$ in each of the $i$ detectors in a network of $D$ detectors, we can write 
\begin{equation}
\label{eq:zn}
Z^N_i = \mathcal{N}(d_i) = \frac{1}{2\pi \sigma^2_i} \ \text{exp}\left(-\frac{1}{2} \frac{d_i}{\sigma^2_i} \right) \ ,
\end{equation}
where $\mathcal{N}(d_i)$ is a normal distribution with a mean of zero and variance proportional to the \psd.

\subsubsection{Coherent Signal Model}
We model coherent signal using a binary black hole waveform template \template, where the vector \parameters contains a point in the 15 dimensional space describing precessing binary-black hole mergers. For the signal to be coherent, \parameters must be consistent in each $4\ \text{s}$ data segment $d_i$ for a network of $D$ detectors, Hence, the coherent signal evidence is calculated as
\begin{equation}
\label{eq:zs}
% Z^S = \int\limits_{\vec{\theta}} \prod\limits^{D}_{i=1} \left[\frac{1}{2\pi \sigma^2_i} \ \text{exp}\left(-\frac{1}{2} \frac{|d_i-\mu(\vec{\theta})|^2}{\sigma^2_i} \right) \right] \pi(\vec{\theta}| \mathcal{H}_S)\  \text{d}\vec{\theta} \ ,
Z^S = \int\limits_{\vec{\theta}} \prod\limits^{D}_{i=1} \left[ \mathcal{L}(d_i|\mu(\vec{\theta}))\right] \pi(\vec{\theta}| \mathcal{H}_S)\  \text{d}\vec{\theta} \ ,
\end{equation}
where $\pi(\theta| \mathcal{H}_S)$ is the prior for the parameters in the coherent signal hypothesis, and $\mathcal{L}(d_i|\mu(\vec{\theta}))$ is the likelihood for the coherent signal hypothesis that depends on the gravitational wave template \template and its parameters \parameters.

\subsubsection{Incoherent Glitch Model}
Finally, as glitches are challenging to model and poorly understood, we utilise a surrogate model for glitches: the glitches are modelled using gravitational wave templates  \template with uncorrelated  parameters amongst the different detectors such that  $\vec{\theta}_i \neq \vec{\theta}_j$ for two detectors $i$ and $j$ \cite{bci}.  Modelling glitches with \template captures the worst case scenario: when glitches appear similar to gravitational wave signals. Thus, we can write $Z^G_i$ as 
\begin{equation}
\label{eq:zg}
% Z^G_i = \int\limits_{\vec{\theta}} \frac{1}{2\pi \sigma^2_i} \ \text{exp}\left(-\frac{1}{2} \frac{|d_i-\mu(\vec{\theta})|^2}{\sigma^2_i} \right) \pi(\vec{\theta}| \mathcal{H}_G)\  \text{d}\vec{\theta}  \ ,
Z^G_i = \int\limits_{\vec{\theta}} \mathcal{L}(d_i|\mu(\vec{\theta}))\ \pi(\vec{\theta}| \mathcal{H}_G)\  \text{d}\vec{\theta}  \ ,
\end{equation}
where $\pi(\theta| \mathcal{H}_G)$ is the prior for the parameters in the incoherent glitch hypothesis.

@@

After calculating the \bcr for a set of background triggers and simulated triggers from as stretch of detector-data, we can compute probability distributions for the background and simulated triggers, $p_b(\text{BCR})$ and $p_s(\text{BCR})$. We expect the background trigger and simulated signal \bcr values to favour the incoherent glitch and the coherent signal hypothesis, respectively. Ideally these distributions that represent two unique populations should be distinctly separate and have no overlap in their \bcr values. The prior odds parameters $\alpha$ and $\beta$ from Eq.~\ref{eq:bcr} help separate the two distributions. Altering $\alpha$ translates the \bcr probability distributions while adjusting $\beta$ spreads the distributions. Although Bayesian hyper-parameter estimation can determine the optimal values for these parameters, an easier approach is to manually adjust the parameters for each data chunk's \rs{have you introduced the idea of the pycbc chunk yet?}  \bcr distribution. In this study, we manually tune $\alpha$ and $\beta$ to maximally separate the \bcr distributions for the background and simulated triggers. \\

To calculate the separation between $p_b(\text{BCR})$ and $p_s(\text{BCR})$, we use the Kullback--Leibler divergence (KL divergence) $D_{KL}$, given by
\begin{equation}
    D_{KL}(p_b | p_s) = \sum\limits_{i=1}^N p_b(\text{BCR}_i) \cdot (\log p_b(\text{BCR}_i) - \log p_s(\text{BCR}_i)) \ .
\end{equation}
The $D_{KL}=0$ when the distributions are identical and increases as the asymmetry between the distributions increases. \\

We limit our search for the maximum KL-divergence in the $\alpha$ and $\beta$ ranges of $[E-10, E0]$ as values outside this range are nonphysical. We set our values for $\alpha$ and $\beta$ to those which provide the highest KL-divergence and can now calculate the \bcr for candidate events present in this stretch of detector-data. Note that we conduct the analysis in stretches of data rather than an entire data set as the background may be different at different points of the entire data set.

\subsection{Calculating the Significance of Candidate events}
With the tuned values of $\alpha$ and $\beta$ we can calculate the \bcr for candidate events. As mentioned previously, irrespective of the \bcr's Bayesian interpretation, we treat the \bcr as a traditional detection statistic to obtain a frequentist estimate of the significance of candidate event measured against the background of signal-free data. \\

>>>>

\documentclass[%
%reprint,
%superscriptaddress,
%groupedaddress,
%unsortedaddress,
%runinaddress,
%frontmatterverbose, 
% preprint,
%preprintnumbers,
%nofootinbib,
%nobibnotes,
%bibnotes,
 amsmath,amssymb,
 aps,
%pra,
%prb,
%rmp,
%prstab,
%prstper,
%floatfix,
twocolumn
]{revtex4}

@@

\newcommand{\bilby}{{\sc \href{https://lscsoft.docs.ligo.org/bilby/}{\texttt{Bilby}}}\xspace}
\newcommand{\bilbypipe}{{\sc bilby\_pipe}\xspace}
\newcommand{\pbilby}{{\sc pBilby}\xspace}
\newcommand{\dynesty}{{\sc dynesty}\xspace}
\newcommand{\cpnest}{{\sc cpnest}\xspace}
\newcommand{\ptemcee}{{\sc ptemcee}\xspace}
\newcommand{\gwpy}{{\sc \href{https://gwpy.github.io/}{\texttt{GWpy}}}\xspace}
\newcommand{\imrphenomp}{{\sc IMRPhenomPv2}\xspace}
\newcommand{\seob}{{\sc SEOBNRv4PHM}\xspace}
\newcommand{\pycbc}{{\sc \href{https://pycbc.org/}{\texttt{PyCBC}}}\xspace}
% MATH COMMANDS 
\newcommand{\mathcmd}[1]{{\sc \relax\ifmmode#1\else $#1$\fi}\xspace}
\newcommand{\bcr}{\mathcmd{\rho_\text{B}}}
\newcommand{\pycbcstat}{\mathcmd{\rho_\text{PC}}}
\newcommand{\snr}{\mathcmd{\rho}}
\newcommand{\psd}{\mathcmd{P(f)}}
\newcommand{\msun}{\mathcmd{\text{M}_\odot}}
\newcommand{\parameters}{\mathcmd{\vec{\theta}}}
\newcommand{\prior}{\mathcmd{\pi(\parameters)}}
\newcommand{\template}{\mathcmd{\mu(\parameters)}}

\newcommand{\pastro}{\relax\ifmmode{p_\text{astro}}\else $p_\text{astro}$\fi\xspace}
\newcommand{\pval}{\relax\ifmmode{\text{p-value}}\else $\text{p-value}$\fi\xspace}

@@

\preprint{\textcolor[rgb]{0.00,0.50,0.75}{{\input{.githash.txt}}}}

\title{A Bayesian search to find \\high-mass black holes in LIGO data}

@@

The detection of intermediate mass black holes ( $10^2-10^6$~\msun) will shed light on the formation of supermassive black holes and thus galaxy formation. Although LIGO is sensitive to the merger of binary black holes with total masses up to $500$~\msun, only 1 of their 50 detections have a total mass $>100$~\msun. A possible explanation for the absence of intermediate mass events may be due to their misclassification as short-duration instrumental noise transients. Short-duration instrumental transients mimic the short-duration gravitational-wave signals from intermediate mass binary black hole mergers. Here we demonstrate that a search method utilising Bayesian inference could be a more sensitive tool for detecting high-mass binary black hole mergers (systems with a total mass $>55$~\msun) as compared to traditional match-filtering. We have applied this technique on the high-mass triggers during LIGO's second observing run to investigate the possibility of discovering new gravitational-wave signals from high mass black hole binaries, and to re-calculate the significance of high-mass candidate events. Although our search does not discover new candidate events, it does alter the significance of candidates identified by various search pipelines.

@@

%%%%%%---SECTIONS---%%%%%%%%%%%%
\section{\label{sec:Introduction}Introduction}
Since the 1970s, there has been an accumulation of evidence for stellar-mass and supermassive black holes. In April 2019, the Event Horizon Telescope provided the first visual evidence of the supermassive black hole M87 \cite{m87photo}. As of November 2020, the LIGO Scientific Collaboration has confirmed $\sim50$ binary black hole systems and listed numerous candidate events \cite{GWTC1, GWTC2, IAS0, IAS1, IAS2, pycbc_ogc_2}. These various discoveries have firmly established the existence of stellar-mass black holes, supermassive black holes and binary black hole systems. Until recently, there was no definitive evidence for intermediate-mass black holes, the black holes that lie in between stellar-mass and supermassive black hole systems with masses between $10^2-10^6\ \msun$. This changed with the detection of GW190521, a unique gravitational wave event that lead to the formation of a black hole with a mass  $142\ \msun$, the first direct discovery of an intermediate mass black hole. Although this is the first gravitational wave that has l;ead to the discovery of a black hole with a mass of $>100\ \msun$, ground based gravitational wave detectors are sensitive to gravitational waves from even more massive systems, up to a systems with a total mass of $400\ \msun$. Gravitation waves from systems with masses $>100\ \msun$ systems should occur at a rate of $0-10\text{yr}^{-1}$ \cite{fregeau2006imbhbRatePrediction, mandel2008rates,rodriguez2015bbhRatePredictions,ligo_imbh_search}.

However, even after conducting a targeted matched-filter based search for gravitational waves from intermediate-mass black holes the largest total mass detected so far is approximately $80\ \msun$ \cite{imbhbSearch2014, ligo_imbh_search, abbott2019gwtc}. A possible explanation for the absence of intermediate-mass events may be due to their misclassification as short-duration instrumental noise transients known as glitches \cite{glitch_in_fifth_ligo_run, bayeswave, improving_dq_in_early_runs, ligo_glitch_gw150914, pycbc_short_duration_transients, pe_with_glitch, blip_glitches}. These glitches can mimic astrophysical signals and hence decrease the significance of true gravitational wave events.

One method to account for glitches while searching for gravitational waves from coalescing compact binaries is by utilising an astrophysical Bayesian odds  \cite{bci, BCR1, BCR2, bcr_gw151216, bayesian_odds}. A true Bayesian odds calculated without using bootstrap techniques can provide events with a more accurate significance that is agnostic to a specific search strategy \cite{BCR2, bcr_gw151216,  bayesian_odds}. Additionally, a Bayesian odds can include more information than is included in current matched filter searches such as if the gravitational wave event's signal is coherent amongst the network of detectors, if the binary system that created the gravitational waves was precessing, and if the gravitational wave signal contains higher-order modes.  It is because Bayesian methods can incorporate all this physical information about a gravitational wave signal that the LIGO Scientific Collaboration uses these methods to determine the source parameters of gravitational wave events \cite{abbott2016ligo, abbott2019gwtc}. This paper demonstrates that the power of Bayesian methods used in parameter estimation can also successfully be used to discriminate between coherent gravitational-wave signals, incoherent glitches, and Gaussian noise in the form of a Bayesian search.

In this paper we utilise a Bayesian method, called the Bayesian Coherence Ratio \bcr \cite{BCR1}, to search for the significant gravitational wave signals from high-mass (systems with total masses in the range of $55-500\ \msun$) compact binary coalescences in the detector data recorded during aLIGO's second observing run (O2).  We find that (a) our search does not identify any unreported stellar mass or intermediate mass black holes; (b) high-mass events reported in the GWTC-1, including GW170729 (an event with disputed \pastro amongst various search pipelines) have high significance; \av{and that (c) high-mass events detected from the IAS groups have low significance. }

@@

\section{Method\label{sec:method}}
The gravitational wave community uses Bayesian inference to perform parameter estimation and model selection. In this work, we utilise Bayesian inference to calculate the significance of high-mass candidate events in O2 by using the \bcr as a ranking statistic, taking a step forward to building a unified Bayesian framework to search for candidates and estimate their parameters.

Although a dedicated Bayesian search for gravitational waves, as presented by ~\cite{BCR2}, does not require noise estimation using empirical methods, this Bayesian significance ranking technique utilises \textit{time-slides}. To perform time-slides, the data from independent observatories are time-shifted by amounts greater than the light-travel time between the two detectors. Each unique time-slide amount creates an artificial signal-free ``background'' data set. When gravitational-wave search pipelines scan these background data sets for gravitational wave events, they can find candidate triggers, even though the background data set should not contain coherent astrophysical signals. These candidate triggers in background data are labelled background triggers, while coherent triggers obtained in non-time-slid data are labelled candidate triggers.

Calculating the \bcr ranking statistic for each background trigger builds a background \bcr distribution. With a background distribution,  it is possible to assign a statistical significance of how likely a candidate trigger, detected by the search pipelines over non-time slid data, is due to a gravitational wave signal.

This section discusses (a) the method to retrieve triggers, and (b) the \bcr and how it is utilised as a ranking statistic to calculate the significance of candidate triggers.

\subsection{Triggers for Analysis}

The LIGO Scientific collaboration operates several search pipelines that scan for gravitational waves from compact binary mergers such as \texttt{GstLAL}, \texttt{MBTA}, \texttt{SPIIR} and \pycbc \cite{abbott2019gwtc}.

The output of \pycbc's search is a list of candidate trigger times and their corresponding \pycbc ranking statistic \pycbcstat. The \pycbcstat statistic is akin to the matched-filter signal-to-noise ratio \snr. However, unlike \snr, \pycbcstat includes some information on the candidate signal's intrinsic and extrinsic properties and other information that feeds into determining if the signal can have astrophysical origins \cite{pycbc_og6}. The additional physical information incorporated in \pycbcstat makes it a more accurate measure of significance than the standard \snr.

Whenever a local maximum of $\pycbcstat > \snr_\text{T}$, where $\snr_\text{T}$ is some threshold value, the search pipeline produces a single-detector trigger associated with the detector and time where the apparent signal in the data has its merger.

For \pycbc to consider a trigger to be a \textbf{candidate trigger}, a trigger from astrophysical origins, the trigger must be observed between detectors with the same template and a time of arrival difference less than the gravitational-wave travel time \cite{pycbc_og6}. To test its search, \pycbc also conducts searches for \textbf{simulated triggers}, artificial triggers manufactured by injecting signals into the detector data. These simulated signal studies provide \pycbc with metrics on its search's sensitivity. Finally, to quantify the statistical significance of candidate events, \pycbc artificially constructs a \textbf{background trigger} set to compare against the candidate events. These background triggers are coherent signal-free events, constructed by applying relative offsets, or time-slides, between the data of different detectors \cite{pycbc_og6}. Note that the time-slides to generate the background triggers are greater than the gravitational-wave travel time between detectors to ensure that the background triggers are not of astrophysical origins. \\

Our work demonstrates that the \bcr can be used in the same way as \pycbcstat to measure candidate triggers' statistical significance.  The \bcr can be a powerful ranking statistic as the \bcr incorporates information of not only all possible binary black hole systems that might have merged to produce the trigger but also the various incoherent glitches that might cause a false-detection.

Before we discuss how we use the \bcr as a measure of significance, we introduce the method to calculate the \bcr in the following section.

@@

Bayes theorem states that the posterior probability distribution $p(\parameters|d,\mathcal{H})$ for data $d$ and a vector of parameters \parameters that describe a model which quantifies a hypothesis $\mathcal{H}$, is given by
\begin{equation}
p(\parameters|d,\mathcal{H}) = \frac{\mathcal{L}(d|\parameters, \mathcal{H}) \ \pi(\parameters | \mathcal{H})}{\mathcal{Z}(d|\mathcal{H})}\ , 
\end{equation}
where $\mathcal{L}(d|\parameters, \mathcal{H})$ is the likelihood of the data given the parameters \parameters and the hypothesis, $\pi(\parameters | \mathcal{H})$ is the prior probability of the parameters, and finally,
\begin{equation}
    \mathcal{Z}(d|\mathcal{H}) = \int\limits_{\parameters} \ \mathcal{L}(d|\parameters,\mathcal{H}) \ \pi(\parameters | \mathcal{H}) \ d\parameters
\end{equation} is the likelihood after marginalising over the parameters \parameters.  To compare two hypotheses $\mathcal{H}_A$ and $\mathcal{H}_B$ with the Bayes theorem one can calculate an odds ratio
\begin{equation}
    \mathcal{O}^A_B = \frac{\mathcal{Z}^A\ \pi(\parameters^A)}{\mathcal{Z}^B\ \pi(\parameters^B)}\ ,
\end{equation}
where $\mathcal{Z}^A$ and $\mathcal{Z}^B$ are the shorthand for the evidences  $\mathcal{Z}(d|\mathcal{H}_A)$ and $\mathcal{Z}(d|\mathcal{H}_B)$. The odds ratio can tell us which of the two hypotheses is more likely. For example, if $\mathcal{O}^A_B >> 1$, then this odds ratio indicates that the $\mathcal{H}_A$ describes the data much better than $\mathcal{H}_B$.

The \bcr is a Bayesian odds ratio like the above, of a coherent signal hypotheses $\mathcal{H}_S$ and an incoherent instrumental feature hypothesis $\mathcal{H}_I$ for a network of $D$ detectors. $\mathcal{H_I}$ states that each detector $i$ has either pure Gaussian noise $\mathcal{H}_N$ or a glitch $\mathcal{H}_G$. Following \citet{BCR1}, the \bcr is given by
\begin{equation}
\label{eq:bcr}
\bcr = \frac{\alpha Z^S}{\prod\limits^D_{i=1} \ [\beta Z^G_i + (1-\beta)Z^N_i]}\ ,
\end{equation}
where $Z^S$, $Z^G_i$ and $Z^N_i$ are the Bayesian evidences (marginalised likelihoods) for $\mathcal{H}_S$, $\mathcal{H}_N$, and $\mathcal{H}_G$. $\alpha$ and $\beta$, are the prior odds for obtaining a signal $\alpha=P(\mathcal{H}_S)/P(\mathcal{H}_I)$ and the prior odds for obtaining a glitch $\beta=P(\mathcal{H}_G)/P(\mathcal{H}_I)$. As the rate of signal and glitches are unknown, these priors $\alpha$ and $\beta$ are tuned to maximise the \bcr distributions for background data (signal-free data) and simulated signals \cite{BCR1}.

\subsection{Bayesian Evidence Evaluation}
\subsubsection{Noise Model}
We assume that each detector's noise is Gaussian and stationary over the period being analysed \cite{ligo_psd}. In practice, we assume that the noise has a mean of zero that the noise variance $\sigma^2$ is proportional to the noise power spectral density (PSD) \psd of the data. Using the \psd, for each data segment $d_i$ in each of the $i$ detectors in a network of $D$ detectors, we can write 
\begin{equation}
\label{eq:zn}
Z^N_i = \mathcal{N}(d_i) = \frac{1}{2\pi \psd_i} \ \text{exp}\left(-\frac{1}{2} \frac{d_i}{\psd_i} \right) \ ,
\end{equation}
where $\mathcal{N}(d_i)$ is a normal distribution with $\mu=0$ and $\sigma^2\sim \psd$.

\subsubsection{Coherent Signal Model}
We model coherent signal using a binary black hole waveform template \template, where the vector \parameters contains a point in the 15 dimensional space describing precessing binary-black hole mergers. For the signal to be coherent, \parameters must be consistent in each $4\ \text{s}$ data segment $d_i$ for a network of $D$ detectors, Hence, the coherent signal evidence is calculated as
\begin{equation}
\label{eq:zs}
Z^S = \int\limits_{\parameters} \prod\limits^{D}_{i=1} \left[ \mathcal{L}(d_i|\template)\right] \pi(\parameters | \mathcal{H}_S)\  \text{d}\parameters \ ,
\end{equation}
where $\pi(\parameters| \mathcal{H}_S)$ is the prior for the parameters in the coherent signal hypothesis, and $\mathcal{L}(d_i|\template))$ is the likelihood for the coherent signal hypothesis that depends on the gravitational wave template \template and its parameters \parameters.

\subsubsection{Incoherent Glitch Model}
Finally, as glitches are challenging to model and poorly understood, we utilise a surrogate model for glitches: the glitches are modelled using gravitational wave templates  \template with uncorrelated  parameters amongst the different detectors such that  $\parameters_i \neq \parameters_j$ for two detectors $i$ and $j$ \cite{bci}.  Modelling glitches with \template captures the worst case scenario: when glitches appear similar to gravitational wave signals. Thus, we can write $Z^G_i$ as 
\begin{equation}
\label{eq:zg}
Z^G_i = \int\limits_{\parameters} \mathcal{L}(d_i|\template))\ \pi(\parameters| \mathcal{H}_G)\  \text{d}\parameters  \ ,
\end{equation}
where $\pi(\theta| \mathcal{H}_G)$ is the prior for the parameters in the incoherent glitch hypothesis.

@@

After calculating the \bcr for a set of background triggers and simulated triggers from as stretch of detector-data (a data chunk), we can compute probability distributions for the background and simulated triggers, $p_b(\bcr)$ and $p_s(\bcr)$. We expect the background trigger and simulated signal \bcr values to favour the incoherent glitch and the coherent signal hypothesis, respectively. Ideally these distributions that represent two unique populations should be distinctly separate and have no overlap in their \bcr values. The prior odds parameters $\alpha$ and $\beta$ from Eq.~\ref{eq:bcr} help separate the two distributions. Altering $\alpha$ translates the \bcr probability distributions while adjusting $\beta$ spreads the distributions. Although Bayesian hyper-parameter estimation can determine the optimal values for $\alpha$ and $\beta$, an easier approach is to adjust the parameters for each data chunk's \bcr distribution. In this study, we tune $\alpha$ and $\beta$ to maximally separate the \bcr distributions for the background and simulated triggers.

To calculate the separation between $p_b(\bcr)$ and $p_s(\bcr)$, we use the Kullback--Leibler divergence (KL divergence) $D_{KL}$, given by
\begin{equation}
    D_{KL}(p_b | p_s) = \sum\limits_{i=1}^N p_b(\bcr^i) \cdot (\log p_b(\bcr^i) - \log p_s(\bcr^i) \ .
\end{equation}
The $D_{KL}=0$ when the distributions are identical and increases as the asymmetry between the distributions increases.

We limit our search for the maximum KL-divergence in the $\alpha$ and $\beta$ ranges of $[E-10, E0]$ as values outside this range are nonphysical. We set our values for $\alpha$ and $\beta$ to those which provide the highest KL-divergence and calculate the \bcr for candidate events present in this data chunk. Note that we conduct the analysis in data chunks of a few days rather than an entire data set of a few months as the background may be different at different points of the entire data set.

\subsection{Calculating the Significance of Candidate events}
With the tuned values of $\alpha$ and $\beta$ we can calculate the \bcr for candidate events. As mentioned previously, irrespective of the \bcr's Bayesian interpretation, we treat the \bcr as a traditional detection statistic to obtain a frequentist estimate of the significance of candidate event measured against the background of signal-free data.
